---
output: html_document
editor_options: 
  chunk_output_type: inline
---
# Extensions to work with GIS software in R


## RQGIS
There's a bit of overhead to install QGIS and configure RQGIS - I use QGIS but have not set up RQGIS myself, I recommend reading the description in [Geocomputation with R](https://geocompr.robinlovelace.net/gis.html#rqgis)

## R-ArcGIS bridge
See [this description](https://www.esri.com/en-us/arcgis/products/r-arcgis-bridge/get-started) describing how to link your R install to the R-ArcGIS bridge

## Accessing Python toolbox using reticulate
I use this all the time! I highly recommend this approach if you want to integrate python workflows and libraries (including arcpy) with R workflows within reproducible R markdown files.
```{r warning=FALSE, error=FALSE, message=FALSE, eval=TRUE}
library(reticulate)
use_condaenv("StreamCat")
```

We can immediately start playing with python within a code block designated as python
```{python}
import pandas as pd

print('hello python')

some_dict = {'a':1, 'b':2, 'c':3}

print(some_dict.keys())
```

Load our gage data in Python...
```{python}
import pandas as pd
gages = pd.read_csv('C:/Users/mweber/GitProjects/Rspatialworkshop/inst/extdata/Gages_flowdata.csv')
gages.head()
gages['STATE'].unique()
PNW_gages = gages[gages['STATE'].isin(['OR','WA','ID'])]
```

### Access Python objects directly from R
Now work with the pandas data directly within R
```{r}
library(sf)
library(ggplot2)

gages <- st_as_sf(py$PNW_gages,coords = c('LON_SITE','LAT_SITE'),crs = 4269) 
gages <- st_transform(gages, crs=5070) #5070 is Albers system in metres
ggplot(gages) + geom_sf()
```

### And share spatial results from Python
You can work with spatial tools in Python and share results with R!
```{python, eval=FALSE}
from rasterstats import zonal_stats
clnp = 'C:/Users/mweber/Temp/CraterLake_tm.shp'
elev = 'C:/Users/mweber/Temp/elevation_tm.tif'

park_elev = zonal_stats(clnp, elev, all_touched=True,geojson_out=True, stats="count mean sum nodata")
geostats = gp.GeoDataFrame.from_features(park_elev)
```

```{r warning=FALSE, error=FALSE, message=FALSE, eval=FALSE}
zonal <- py$geostats
```

## R Whitebox Tools
We won't go into here but worth mentioning as a rich set of tools you can access in R - [whiteboxR](https://giswqs.github.io/whiteboxR/)

## rgee
I was lucky enough to attend the Google Earth Engine summit pre-pandemic and came away wanting to write [this] exact package](https://github.com/r-spatial/rgee) - someone beat me to it and it's really slick.

Here I'm just running the demo code in the ReadMe for the `rgee` package as a proof of concept of cool things you can do being able to leverage Earth Engine directly in R.  Note that there is overhead in getting this all set up.

```{r warning=FALSE, error=FALSE, message=FALSE, eval=FALSE}
library(reticulate)
library(rgee)
ee_Initialize()
# gm <- import("geemap")
```

Function to create a time band containing image date as years since 1991.
```{r warning=FALSE, error=FALSE, message=FALSE, eval=FALSE}
createTimeBand <-function(img) {
  year <- ee$Date(img$get('system:time_start'))$get('year')$subtract(1991L)
  ee$Image(year)$byte()$addBands(img)
}
```

Using Earth Engine syntax, we 'Map' the time band creation helper over the night-time lights collection.
```{r warning=FALSE, error=FALSE, message=FALSE, eval=FALSE}
collection <- ee$
  ImageCollection('NOAA/DMSP-OLS/NIGHTTIME_LIGHTS')$
  select('stable_lights')$
  map(createTimeBand)
```

We compute a linear fit over the series of values at each pixel, visualizing the y-intercept in green, and positive/negative slopes as red/blue.
```{r warning=FALSE, error=FALSE, message=FALSE, eval=FALSE}
col_reduce <- collection$reduce(ee$Reducer$linearFit())
col_reduce <- col_reduce$addBands(
  col_reduce$select('scale'))
ee_print(col_reduce)
```

We make an interactive visualization - pretty cool!
```{r warning=FALSE, error=FALSE, message=FALSE, eval=FALSE}
Map$setCenter(9.08203, 47.39835, 3)
Map$addLayer(
  eeObject = col_reduce,
  visParams = list(
    bands = c("scale", "offset", "scale"),
    min = 0,
    max = c(0.18, 20, -0.18)
  ),
  name = "stable lights trend"
)

```
